Regional Inequalities and economic growth; cross section and times series analysis

### 2023-12-06 MSB104_Submission_Group1.qmd
mmm
hey
.

The sructure of our study is methodically designed to facilitate this exploration. In part 1A and B we calculate GDP per capita and explore regional inequality using various descriptive statistics and visualizations such as panel data. In part 2A and B we use cross sectional method to examine the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010. Furthermore, exploring alternative functional forms and implementing panel estimation to analzise the relationship between regional development inequality.

@@

Why some places grow and prosper is a fundamental question in the field of social science. This question motivated Adam Smith's \[1863\] work and has been a major influence of the *new economic geography* [@feldman2014]*. Furthermore,* regional growth and inequality have been important topics of research in Europe for many years. Despite the European commission\'s efforts to decrease income gaps, significant disparities in income and economic development persist across regions in Europe [@lessmann2017].

@@

Regional growth in Europe reveals a complex pattern, influenced significantly by factors such as geography, human capital, and economic policies. Wealthier nations like Denmark tend to experience quicker regional convergence, a phenomenon supported by \[@gennaioli2014\] findings.

@@

The 2008 financial crisis disrupted regional growth and worsened income inequalities across Europe. Nguyen et al. (2022) highlight that financial instability typically leads to increased income disparity; a trend observed globally. Focusing on Europe, we expect varied regional economic performances, especially when comparing Eastern European countries such as Hungary and Slovakia to their Western counterparts. Given the role of capital market regulations in promoting regional convergence, outlined by Gennaioli et al (2014), we anticipate a potential growth acceleration after submission.  EU membership provided Slovakia and Hungary with resources for economic recovery, while Western European countries, including Denmark, Portugal, and France, likely experienced significant economic declines due to the financial crisis. This pattern possibly mirrored in the recent COVID-19 pandemic. These complexities highlight the importance of a detailed analysis to understand regional economic growth and inequality in challenging times.

@@

The effect that surprised us most was that the economic crisis led to increase inequality\[@nguyen2022\]. indicate that any type of financial crisis results in higher income inequality. Furthermore,  @lessmann2017 address the importance of studying these aspects. Moreover, its potential consequences, such as political tensions that can undermine social and political stability [@lessmann2017].

@@

Geographic characteristics stand out as fundamental, with the natural and topographical elements of a region shifting income distributions [@lessmann2017]. Urbanization emerges as a crucial determinant, where active, populations around urban centres usually enjoy raised income levels, contrasting with the disparities witnessed in more rural and secluded areas [@lessmann2017].

@@

[@lessmann2017] also discusses the determinant of Education in the context of human capital. The study mentions that human capital is the most important determinant of differences in regional development within countries. Moreover, the quality of human capital within countries, as measured by the secondary-school enrolment rate, can decrease regional inequality by facilitating regional spillovers and convergence. Lastly, promoting internal migration.

@@

@lessmann2017 provides insights into the inverted U-shaped relationship between regional inequality and the level of economic development in different country groups [@lessmann2017]. The inverted U-shape also referred to as Kuznets Curve suggests that as an economy develops, regional inequality initially rises, reaches a peak, and then starts to decline. Additionally, there is an N-shaped relationship between regional inequality and economic development according to \@lessmann2017. Subsequently, regional inequality increases again after the inverted U-shaped pattern has been completed [@lessmann2017].

Lastly, [@iammarino2019] argue that there are some countries in the EU that are more evenly developed than others, and that a map of underperformance or over-performance means less in a high-income but evenly developed country, such as in Denmark in our case [@iammarino2019]. On the other hand, France has been a subject of discussion and policy making for many years, regarding its well-known pockets of extreme wealth in regions such as Cap d'Antibes.

Overall, it is evident that regional economic growth and inequality in Europe from 2000 to 2020 have been influenced by several factors, with transport infrastructure and education emerging as significant determinants. Earlier studies provide us with a fundamental understanding of these dynamics, while also highlighting the impacts of economic crises on regional disparities. The integration of transport infrastructure and education adds an additional layer of complexity, underscoring its potential as a catalyst for regional development and equality.

# Data and Methodology

@@

# Creating a box plot
ggplot(gini_filtered, aes(x = id_nuts0, y = GDP_per_capita)) +
  geom_boxplot() +
  ggtitle("Plot 1: Distribution of GDP per Capita by Country") +
  xlab("Country") +
  ylab("GDP per Capita")

# Saving the plot for referencing in the appendix
g <- ggplot(gini_filtered, aes(x = id_nuts0, y = GDP_per_capita)) +
  geom_boxplot() +
  ggtitle("Distribution of GDP per Capita by Country") +
  xlab("Country") +
  ylab("GDP per Capita")

# Save the plot as an image file
ggsave("GDP_per_Capita_Boxplot.png", plot = g, width = 10, height = 6)
```

The box plots (in the appendix) provide a distinct perspective on the data, independent of the assumptions typically associated with other statistical distributions. This method of visualization highlights Denmark as having the highest overall GDP per capita across its regions. France, while not having the overall highest GDP per capita, demonstrates significant variability, especially when considering its outliers. Hungary, in contrast, is still presenting the lowest GDP per capita. These insights, derived from the box plot analysis, augment our understanding from previous descriptive statistics, bar plots, and time series analyses, offering a more nuanced view of the GDP per capita distribution among these countries.

# Empirical Findings

## Cross sectional estimates

@@

@@

The goodness of fit in simple linear regression, evaluated by R\^2, gauges how well the regression line fits the data. This line is determined using the least squares criterion, minimizing squared differences between data points and the line. R\^2 varies from 0 to 1, where 0 indicates no explanatory power and 1 signifies perfect fit. A higher R\^2 suggests a better model fit. For an effective model, data should linearly align with minimal patterns in residuals, assuming all linear regression prerequisites are met.

**Residuals vs. Fitted Values Plot** is used to check for homoscedasticity and linearity. Ideally, this plot shows no pattern; the residuals are randomly scattered around the horizontal line at zero. If there's a pattern (like a curve or systematic spread of residuals), it suggests non-linearity or heteroscedasticity.

@@

# Made a copy for appendix
# Finding the natural logarithm of GDP per capita
data2010 <- data2010 |> 
  mutate(log_GDP_per_capita = log(GDP_per_capita))

@@

```{r}
#| echo: false
#| output: false
#| fig-cap: Regression statistics of all countries seperately for the year 2010

#Appendix 
# Estimation for each country seperately 
lmSK <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "SK"))
lmDK <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "DK"))
lmHU <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "HU"))
lmPT <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "PT"))
lmFR <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "FR"))

@@

The OLS method is according to @wooldridge2020 employed to identify the best-fitting linear relationship between the dependent and independent variables, aiming to minimize the sum of squared residuals. Furthermore, this technique ensures that the estimations of the intercept (β0) and slope (β1) yield the least possible cumulative discrepancy between the actual and predicted values. The strength of OLS lies in its closed-form solution, providing a straightforward computation of coefficients directly from the dataset, without necessitating iterative procedures. According to @wooldridge2020 the following assumtions need to be met to achieve BLUE.

1\. ***Linear in coefficients and error term***: The model is linear in its terms and the error term.

2\. ***Error term has zero mean***: The error term averages zero, ensuring unbiasedness.

3\. ***No correlation between independent variables and error term***: Independent variables should not predict the error term.

4\. ***Observations of error term are uncorrelated*****:** Each error term observation doesn't predict the next.

5\. ***Homoscedasticity*****:** The error term's variance is consistent across observations.

6\. ***No perfect multicollinearity***: No independent variable is a perfect linear function of others.

7\. ***Normally distributed error term (Optional*****)**: This aids in statistical testing, although not required for unbiased OLS estimates

# Making two plots, first one is a residuals vs fitted plot, while the other is a normal Q-Q residuals plot
#plot(lm, which = 1:2) 
```

The residuals vs fitted values and the Q-plot, combined with the regression statistics may help us find out if classical OLS assumptions are met. The first help us check for homoscedasticity assumption of a linear regression model. The residuals should be randomly scattered around the horizontal 0 line, something that can indicate that the variances of the error terms are constant. In our plot, the residuals are in some extent randomly distributed, and there is also no clear pattern; this suggest that there is likely no significant issues with heteroscedasticity or non-linearity. The latter is used to assess if the residuals of the linear model are normally distributed. In our plot, most of the points follow the line closely, suggesting that the residuals are normally distributed. There are however some outliers in the tails, that suggest some variation from normality. Both plots are presented in appendix.

```{r}
#| echo: false
#| output: false
#| label: fig_scatterplot-all
#| cap-location: bottom
#| fig-cap: "Relationship Between GDP per Capita and Gini"

@@

### Part 2B: Exploring Other Determinants of Inequity

#### I. Data Acquisition

@@

#### II. Multiple Linear Regression Model

@@

```{r}
#| echo: false
#| output: false
#| fig-cap: Multiple Linear Regression Model

@@

#### II. Model specification

@@

#### III. Model Interpretation

@@

## Alternative Functional Forms and Panel Estimates

### Testing Development Effects Across Subsets

#### Subset Analysis

@@

#### Subset Analysis Discussion

@@

#### Results Interpretation

@@

### Heteroskedasticity Testing and Causality Discussion

#### Heteroskedasticity Testing

Heteroskedasticity is according to @wooldridge2020 an important factor to consider in econometrics and statistical modelling, especially in the context of regression analysis. Heteroskedasticity refers to the condition in regression analysis where the variance of the error terms (residuals) is not constant across all levels of the independent variables \[@wooldridge2020\]. In simpler terms, the spread of the residuals varies at different points in the regression model. Moreover, heteroskedasticity can lead to inefficient estimators and unreliable hypothesis tests. Leading to the assumption of homoscedastic in MRL not being met. In other words the variance of the error terms is not constant across all levels if the independent variables when heteroskedasticity is present.

@@

# Discussion

# Limitations and Future Research

# Conclusion

@@

1.  **Descriptive statistics**

@@

2.  **Boxplot**

![](GDP_pc_Boxplot.png)

3.  **Regression statistics of all countries separately for the year 2010**

cat("Intercept (Coefficient 1):", b0_earn, "\n")
cat("Slope of log GDP per capita (Coefficient 2):", b1_earn, "\n")
```

![](images/Regression_all_countries2010-01.png)

**Testing for OLS assumtions**

```{r}
#| echo: false
#|

>>>>

The sructure of our study is methodically designed to facilitate this exploration. In part 1A and B we calculate GDP per capita and explore regional inequality using various descriptive statistics and visualizations such as panel data (***Assignment_1***). In part 2A and B we use cross sectional method to examine the relationship between regional development and inequality, employing cross sectional estimation techniques for the year 2010 (***Assignment_2***). Furthermore, exploring alternative functional forms and implementing panel estimation to analzise the relationship between regional development inequality (***Assignment_3***).

@@

Why some places grow and prosper is a fundamental question in social science research. This question motivated Adam Smith's work, *An Inquiry into the Nature and Causes of the Wealth of Nations,* (Smith [1863](https://link-springer-com.galanga.hvl.no/article/10.1007/s11187-014-9574-4#ref-CR32 "Smith, A. (1863). An inquiry into the nature and causes of the wealth of nations. Edinburgh: A. & C Black.")) and has been a major influence of the *new economic geography* [@feldman2014]*. Furthermore,* regional growth and inequality have been important topics of research in Europe for many years. In recent decades, the European Union implemented policies aimed at reducing regional disparities and promoting economic growth across its member states. Furthermore, despite these efforts, significant disparities in income and economic development persist across regions in Europe [@lessmann2017].

@@

Regional growth in Europe exhibits a complex pattern, influenced significantly by factors such as geography, human capital, and economic policies. Lessmann and Seidel's research offers insights into these dynamics, highlighting the persistence of income disparities, across regions despite overarching economic advancement [@lessmann2017]. Wealthier nations like Denmark tend to experience quicker regional convergence, a phenomenon supported by Gennaioli et al.'s findings [@gennaioli2014].

@@

The 2008 financial crisis significantly disrupted regional growth and exacerbated income inequalities across Europe. Nguyen et al. highlight that financial turmoil typically leads to increased income disparity; a trend observed globally [@nguyen2022].

Focusing on Europe, we expect varied regional economic performances, especially when comparing Eastern European countries such as Hungary and Slovakia to their Western counterparts. Given the role of capital market regulations in promoting regional convergence, outlined by Gennaioli et al., we anticipate a potential growth acceleration. In particular for Eastern European countries following their EU accession in 2003 and 2004 [@gennaioli2014]. However, the financial crisis and subsequent labour market shifts may also have influenced these trends.

EU membership provided Slovakia and Hungary with resources for economic recovery, while Western European countries, including Denmark, Portugal, and France, likely experienced significant economic declines due to the financial crisis, a pattern possibly mirrored in the recent COVID-19 pandemic. In particular Portugal as the crisis hit them harder. These complexities highlight the importance of a detailed analysis to understand regional economic growth and inequality in challenging times.

@@

Another affect that surprised us was that the economic crisis led to increase inequality. The study "*the effects of financial crisis on regional inequality*" indicate any type of financial crisis results in higher income inequality [@nguyen2022].

@lessmann2017 regional inequality study address the importance of studying these aspects. Its potential consequences, such as political tensions that can undermine social and political stability [@lessmann2017]. The paper also discusses the relationship between regional inequality and personal income inequality and conflict risk.

@@

Geographic characteristics stand out as pivotal, with the natural and topographical elements of a region significantly swaying income distributions [@lessmann2017].

Urbanization emerges as a crucial determinant, where bustling, populous around urban centers usually enjoy elevated income levels, contrasting starkly with the often-stark disparities witnessed in more rural and secluded areas [@lessmann2017].

*Transport infrastructure and Education*

@@

[@lessmann2017] also discusses the determinant of Education in the context of human capital. The study mentions that human capital is the most important determinant of differences in regional development within countries. The study also notes that human capital creates positive externalizes, which can spill over regional boundaries. Moreover, the quality of human capital within countries, as measured by the secondary-school enrollment rate, can decrease regional inequality by facilitating regional spillovers and convergence. Lastly, promoting internal migration.

@@

Furthermore, @lessmann2017 provides us with insights into the inverted U-shaped relationship between regional inequality and the level of economic development in different country groups [@lessmann2017]. The inverted U-shape also referred to as Kuznets Curve suggests that as an economy develops, regional inequality initially rises, reaches a peak, and then starts to decline.

@lessmann2017 paper also finds that there is an N-shaped relationship between regional inequality and economic development, which means that regional inequality increases again after the inverted U-shaped pattern has been completed [@lessmann2017].

@iammarino2019 discusses regional inequalities in European countries. It presents evidence that inter-regional inequality has been increasing in the European Union since the turn of the millennium. The authors argue that this is due to the existence of several groups of regional economies in Europe that are structurally very different from one another [@iammarino2019]. Additionally, the authors argue that there are some countries in the EU that are more evenly developed than others, and that a map of under performance or over-performance means less in a high-income but evenly developed country, such as in Denmark in our case [@iammarino2019]. On the other hand, France has been a subject of discussion and policy making for many years, regarding its well-known pockets of extreme wealth in regions such as Cap d'Antibes.

Overall, it is evident that regional economic growth and inequality in Europe from 2000 to 2020 have been influenced by a constellation of factors, with transport infrastructure and education emerging as significant determinants. Earlier studies provide us with a fundamental understanding of these dynamics, while also highlighting the nuanced impacts of economic crises on regional disparities. The integration of transport infrastructure and education into this discourse adds an additional layer of complexity, underscoring its potential as a catalyst for regional development and equality.

# Data and Descriptive

@@

# Creating a box plot
ggplot(gini_filtered, aes(x = id_nuts0, y = GDP_per_capita)) +
  geom_boxplot() +
  ggtitle("Distribution of GDP per Capita by Country") +
  xlab("Country") +
  ylab("GDP per Capita")
```

# Cross sectional estimates

@@

### **Data acquisition**

Firstly, the data is summerised for the year 2010 at the NUTS2 regional level, focusing on economic metrics. Processing the dataset containing the NUTS3 level. Thereafter, computing the total GDP and total population for each NUTS2 region and year. Then, derive the GDP per capita at the NUTS2 level by dividing the total GDP by the total population again. Furthermore, narrowing down the dataset to observations from the year 2010 where the Gini coefficient is positive. Then adding, logarithmic transformations to linearize relationships or to reduce the impact of extreme values.

The next step is making a data frame of NUTS2 2010 and grouping by NUTS0 country level variables, as well as allocating NUTS2 to selected countries. Lastly presenting data in **table XXX** showing a count of NUTS2 regions within each top-level region or country in the 2010 data.

The next step is, estimating a Linear Regression for All Countries using cross sectional data from 2010, NUTS2.

@@

The goodness of fit in a simple linear regression model measures how well the regression line approximates the real data points. The regression line that best represents the data according to the least squares criterion, which minimizes the sum of the squared vertical distances of the points from the line.

R\^2 is a statistical measure that represents the proportion of the variance for the dependent variable that's explained by the independent variable. It ranges from 0 to 1. A higher R\^2 value indicates a better fit of the model to the data.

R\^2 = 0 The model does not explain any of the variability of the response data around its mean.

R\^2 = 1 The model explains all the variability of the response data around its mean.

If the assumptions for a simple linear regression are met, it indicates a good fit for the model. [Plotting the data can provide a visual indication of the goodness of fit.]{.underline} The points should fall around a straight line without clear patterns in the residuals. As it is crucial to verify these assumptions before proceeding with interpreting the results of the regression analysis.

**Residuals vs. Fitted Values Plot** to check for homoscedasticity and linearity. Ideally, this plot shows no pattern; the residuals are randomly scattered around the horizontal line at zero. If there's a pattern (like a curve or systematic spread of residuals), it suggests non-linearity or heteroscedasticity.

@@

# Finding the natural logarithm of GDP per capita
data2010 <- data2010 |> 
  mutate(log_GDP_per_capita = log(GDP_per_capita))

@@

```{r}
#| echo: false
#| fig-cap: Regression statistics of all countries seperately for the year 2010

# Estimation for each country seperately 
lmSK <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "SK"))
lmDK <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "DK"))
lmHU <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "HU"))
lmPT <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "PT"))
lmFR <- lm(gini_n2 ~ log_GDP_per_capita, subset(data2010, id_nuts0 == "FR"))

@@

"The OLS method is employed to identify the best-fitting linear relationship between the dependent and independent variables, aiming to minimize the sum of squared residuals. This technique ensures that the estimations of the intercept (β0) and slope (β1) yield the least possible cumulative discrepancy between the actual and predicted values. The strength of OLS lies in its closed-form solution, providing a straightforward computation of coefficients directly from the data-set, without necessitating iterative procedures.

Furthermore, when the classical linear regression assumptions are met, OLS guarantees that the estimators are BLUE, ensuring their unbiasedness and efficiency. This is particularly crucial in econometric analysis, where the precision and reliability of parameter estimates are paramount for policy implications and economic interpretations. The foundation assumptions of linear regression, including linearity, independence, homoscedasticity, and normality of residuals, are prerequisites for OLS to attain these desirable properties. It is imperative, therefore, to conduct diagnostic tests and assess the validity of these assumptions to ensure the robustness of the OLS estimators used in our analysis."

### **Assumptions**

**I. Linearity**

The relationship between the independent variable X and the dependent variable Y is linear. This means that changes in X are associated with proportional changes in Y. A straight line should provide a good fit to the data points when plotted on a graph.

You can create a scatter plot of Y versus X and visually inspect whether a straight line could well represent the relationship. Alternatively, you can plot the residuals versus the fitted values and check for any obvious patterns or non-linearity.

**II. Independence:**

Observations are independent of each other. The value of Y for one observation should not depend on the value of Y for any other observation.

This assumption is more about study design and data collection. Ensure that your observations are not correlated in time or space. For example, if you're analyzing economic data across regions, make sure that the regions are not influencing each other.

**III. Homoscedasticity (Equal Variance):**

The variance of the residuals (the errors) is constant across all levels of X. This means that the spread or "width" of the residuals should remain roughly the same across all values of the independent variable.

A plot of residuals versus fitted values should show a random scatter and not display any funnel-like shapes (wider at one end).

**IV. Normality of Residuals:**

The residuals (the differences between observed and predicted values) are normally distributed. This assumption is particularly important for hypothesis testing and creating confidence intervals.

A Quantile-Quantile (Q-Q) plot of the residuals can show if they follow a normal distribution. The points should fall roughly along a straight line.

**V. No Perfect Multicollinearity (Specific to Multiple Regression)**

In multiple regression settings, this assumes that no independent variable is a perfect linear function of any other independent variables. While this is more pertinent to multiple linear regression, it is crucial there because high correlation between independent variables can lead to unstable coefficient estimates.

Checking the variance inflation factor (VIF) for each variable; a VIF above 5-10 indicates a problematic amount of collinearity.

**Visualization**

We've made two plots that can help us understand the relationship between GDP per capita and the Gini coefficient, and that together with the regression statistics can help with discussing if the classical OLS assumptions hold for the model. The first one is a plot that shows us residuals vs fitted values, which can help us check the homoscedasticity assumption of a linear regression model. The residuals should be randomly scattered around the horisontal 0 line, something that can inducate that the variances of the error terms are constant. In our plot, the residuals are in some extent randomly distributed, and there is also no clear pattern; this suggest that there is likely no significant issues with heteroscedasticity or non-linearity.

The other plot - normal Q-Q plot - is used to assess if the residuals of the linear model are normally distributed. In our plot, most of the points follow the line closely, suggesting that the residuals are normally distributed. There are however some outliers in the tails, that suggest some variation from normality.

In both of these plots, we can see that there are some outliers, these can affect the fit of the model. These outliers might come from regions that have a high GDP per capita compared to rest of the regions, like Paris in France that is a financial hub.

```{r}
#| echo: false
#| label: fig_scatterplot-all
#| cap-location: bottom
#| fig-cap: "Relationship Between GDP per Capita and Gini"

@@

## Part 2B: Exploring Other Determinants of Inequity

### I. Data Acquisition

@@

### II. Multiple Linear Regression Model

@@

```{r}
#| echo: false
#| fig-cap: Multiple Linear Regression Model

@@

### Model specification

@@

### III. Model Interpretation

@@

# Assignment 3: Alternative Functional Forms and Panel Estimates

## Testing Development Effects Across Subsets

### Subset Analysis

@@

### Subset Analysis Discussion

@@

### Results Interpretation

@@

## Heteroskedasticity Testing and Causality Discussion

### Heteroskedasticity Testing

Heteroskedasticity is according to @wooldridge2020 an important factor to consider in econometrics and statistical modelling, especially in the context of regression analysis. Heteroskedasticity refers to the condition in regression analysis where the variance of the error terms (residuals) is not constant across all levels of the independent variables [@wooldridge2020]. In simpler terms, the spread of the residuals varies at different points in the regression model. Moreover, heteroskedasticity can lead to inefficient estimators and unreliable hypothesis tests. Leading to the assumption of homoscedastic in MRL not being met. In other words the variance of the error terms is not constant across all levels if the independent variables when heteroskedasticity is present.

@@

@@

@@
